---
title: "Untitled"
output: pdf_document
---

```{r, warning=FALSE, message=FALSE}
library(mosaic)
library(readr)
library(tidyverse)
library(broom)
library(survival)
library(survminer)
library(praise)
library(boot)
```

```{r, warning=FALSE, message=FALSE}
df <- read_csv("HELPdata.csv", na="*")
```

```{r, message=FALSE, warning=FALSE}
df <- df %>%
  mutate(yrs_education = as.numeric(a9), 
         gender=a1, 
         alcq_30 = as.numeric(alcq_30), 
         marriage = as.factor(a10), 
         employment = as.factor(a13), 
         income = as.factor(case_when(a18 == 1 ~ "<5000", 
                                      a18 == 2 ~ "5000-10000", 
                                      a18 == 3 ~ "11000-19000",
                                      a18 == 4 ~ "20000-29000",
                                      a18 == 5 ~ "30000-39000",
                                      a18 == 6 ~ "40000-49000",
                                      a18 == 7 ~ "50000+")),
         income_1yr = as.factor(case_when(a18_rec1 == 0 ~ "$19,000", 
                                      a18_rec1 == 1 ~ "$20,000-$49,000", 
                                      a18_rec1 == 2 ~ "$50,000")), 
         any_util = as.factor(case_when(any_util == 0 ~ "No", any_util == 1 ~ "Yes")), 
         attempted_suicide = as.factor(case_when(g1c == 0 ~ "No", g1c == 1 ~ "Yes")), 
         employment = as.factor(
           case_when(a13 == 1 ~ "Full time", 
                     a13 == 2 ~ "Part time",
                     a13 == 3 ~ "Student", 
                     a13 == 4 ~ "Unemployed",
                     a13 == 5 ~ "Ctrl_envir")), 
         homeless = as.factor(case_when(homeless == 0 ~ "No", homeless == 1 ~ "Yes")),
         hs_grad = as.factor(case_when(hs_grad == 0 ~ "No", hs_grad == 1 ~ "Yes")),
         group = as.factor(case_when(group == 0 ~ "Control", group == 1 ~ "Clinic")),
         # linkstatus = as.factor(case_when(linkstatus == 0 ~ "Did not link to primary care", linkstatus == 1 ~ "Linked to Primary Care")),
         alcohol = as.factor(case_when(alcohol == 0 ~ "Not First Drug", alcohol == 1 ~ "First Drug Alcohol")),
         money_spent_on_alcohol = as.numeric(h16a),
         mh_index = as.numeric(mh),
         num_med_problems = as.numeric(d3),
         num_hospitilizations = as.numeric(d1),
         bothered_by_med = as.factor(case_when(d4 == 0 ~ "Not at all",
                                               d4 == 1 ~ "Slightly",
                                               d4 == 2 ~ "Moderately",
                                               d4 == 3 ~ "Considerably",
                                               d4 == 4 ~ "Extremely")),
         bothered = as.factor(case_when(d4_rec == 0 ~ "No",
                                        d4_rec == 1 ~ "Yes"))) %>% 
  select(group, dayslink, linkstatus, yrs_education, gender, age, alcohol, alcq_30, marriage, employment, income, income_1yr, any_util, attempted_suicide, homeless, hs_grad, money_spent_on_alcohol, mh_index, num_med_problems, num_hospitilizations,bothered_by_med, bothered)
```


Bootstrapping the survival model (what are the assumptions?  what do you conclude?)

## Bootstrapping the Survival Model

Bootstraping is a nonparametric technique that resamples the sampled data and samples from the resampled data. It is most useful for building confidence intervals without the assumption of the central limit theory. However, one important assumption is that the observations were randomly sampled. This holds true for this study as the paper mentions 235 observations were randomized into the help clinic. From bootstrapping we hope to estimate the true parameter of the population, which are the coefficients to our proportional hazard model. 

We are interested in bootstrapping the survival model because the dataset is right censored. The variable, Linkstatus, indicates whether an observation connected to primary care within 12 months. We right censor when an individual did not connect to primary care by 12 months. Given that we have censored data offers multiple resampling options. In bootstrapping, we care about the resampling method because it will directly affect our statisical inference. We will be building a survival model out of resampled data. The library "censboot" will help facilitate the various resampling algorithms. Censboot references "Bootstrap Methods and their Applications" by Davison and Hinkley (1997) for their resampling options. There are four different simulation options we will explore. They are ordinary, model, weird, and conditional. The censboot function requires some statistic to be returned. Since we are comparing multiple sampling methods, we will estimate the surval model coefficients and the loglik deviations from the model. .  We can calculate the t-staistic from the loglik deviations and then find the p-value for the Likelihood Ratio test. The null hypothesis is $H_0: \beta_1 = \beta_2 = \beta_3 = \beta_4 = 0$. The alternative hypothesis is $H_A: \text{ at least one coefficient is not zero}$. 

### Ordinary Resampling

As the name suggests, Ordinary resampling is sampling under a random distribution. Ordinary resapmling is appropriate when the dataset is "subject to random censorship" (Davison et al.) According to the paper we are analysis, the patients were enrolled in a randomized control trial. So if we were to use ordinary resampling, the samping distribution would be equivalent to the original data set. Below 

```{r}
cox.fun <- function(data) {
     data <- data %>% drop_na()
     model <- coxph(Surv(dayslink, linkstatus) ~ group + age + alcohol + yrs_education, data = data)
     out <- c(model$coefficients, model$loglik)
     out
}
model1_data <- df %>% select(group, age, alcohol, yrs_education, dayslink, linkstatus) %>% drop_na()
censboot(data = model1_data, cox.fun , R = 499, sim = "ordinary")
```

### Conditional Bootstrap

Conditional resampling works under the assumption that the censoring variable is unrelated to the response variable. Since we do not have some distribution to sample from, "simulations should be conditional" on the censorship pattern (Davison et al.) The pattern of the censorship can be estimated by providing the original censored data and the reversed censored data. To conditionally sample from the censorship pattern, censboot estimates failure times, which is the time an observation dies off. Each observation is censored or not censored depending on its failure time and the censor distribution. 

```{r}
cond.fun <- function(data) {
  cox <- coxph(Surv(dayslink, linkstatus) ~ group + age + alcohol + yrs_education, data = data)
  c(cox$coefficients, cox$loglik)
}
df <- df %>% select(dayslink, linkstatus, group, age, alcohol, yrs_education) %>% mutate(group = as.numeric(group)) %>% drop_na()
df <- df[order(df$group),]
df <- as.data.frame(df)
s1 <- survfit(Surv(dayslink, linkstatus) ~ group, data = df)
s2 <- survfit(Surv(dayslink-0.001*linkstatus, 1-linkstatus) ~ 1, data = df)
censboot(df, cond.fun, R = 499, strata = df$group,
     F.surv = s1, G.surv = s2, sim = "cond")
```


### Weird Bootstap

The weird samlping method treats each death time as independent binomial variables. Each binomial variable as a probability of $\frac{d_j}{n-j+1}$ probability of dying (DAvison et al.) One key advantage to using the weird resampling method is that it does not need any assumptions about the censor data.

```{r}

```


```{r}
final_model <- coxph(Surv(dayslink, linkstatus) ~ group + age + alcohol + yrs_education, data=df) 
final_model
```


```{r}
# NOT RUN {
library(survival)
# Example 3.9 of Davison and Hinkley (1997) does a bootstrap on some
# remission times for patients with a type of leukaemia.  The patients
# were divided into those who received maintenance chemotherapy and 
# those who did not.  Here we are interested in the median remission 
# time for the two groups.
data(aml, package = "boot") # not the version in survival.
aml.fun <- function(data) {
     # surv <- survfit(Surv(time, cens) ~ group, data = data)
     # out <- NULL
     # st <- 1
     # for (s in 1:length(surv$strata)) {
     #      inds <- st:(st + surv$strata[s]-1)
     #      md <- min(surv$time[inds[1-surv$surv[inds] >= 0.5]])
     #      st <- st + surv$strata[s]
     #      out <- c(out, md)
     # }
     # out
  cox <- coxph(Surv(time, cens) ~ group, data = data)
  cox$loglik
}

ggsurvplot(survfit(Surv(time, cens) ~ group, data = aml))
aml.case <- censboot(aml, aml.fun, R = 499, strata = aml$group)
# Now we will look at the same statistic using the conditional 
# bootstrap and the weird bootstrap.  For the conditional bootstrap 
# the survival distribution is stratified but the censoring 
# distribution is not. 

aml.s1 <- survfit(Surv(time, cens) ~ group, data = aml)
aml.s2 <- survfit(Surv(time-0.001*cens, 1-cens) ~ 1, data = aml)
aml.cond <- censboot(aml, aml.fun, R = 499, strata = aml$group,
     F.surv = aml.s1, G.surv = aml.s2, sim = "cond")


# For the weird bootstrap we must redefine our function slightly since
# the data will not contain the group number.
aml.fun1 <- function(data, str) {
     surv <- survfit(Surv(data[, 1], data[, 2]) ~ str)
     out <- NULL
     st <- 1
     for (s in 1:length(surv$strata)) {
          inds <- st:(st + surv$strata[s] - 1)
          md <- min(surv$time[inds[1-surv$surv[inds] >= 0.5]])
          st <- st + surv$strata[s]
          out <- c(out, md)
     }
     out
}
aml.wei <- censboot(cbind(aml$time, aml$cens), aml.fun1, R = 499,
     strata = aml$group,  F.surv = aml.s1, sim = "weird")

aml
df <- df[order(df$group),]
df
aml
summary(df)
summary(aml)
```

