---
title: "CoxPHModel"
output: pdf_document
---

```{r, warning=FALSE, message=FALSE}
library(mosaic)
library(readr)
library(tidyverse)
library(broom)
library(survival)
library(survminer)
library(praise)
```


```{r, warning=FALSE, message=FALSE}
df <- read_csv("HELPdata.csv", na="*")
```

```{r, message=FALSE}
df <- df %>%
  mutate(yrs_education = as.numeric(a9), 
         gender=a1, 
         alcq_30 = as.numeric(alcq_30), 
         marriage = as.factor(a10), 
         employment = as.factor(a13), 
         income = as.factor(case_when(a18 == 1 ~ "<5000", 
                                      a18 == 2 ~ "5000-10000", 
                                      a18 == 3 ~ "11000-19000",
                                      a18 == 4 ~ "20000-29000",
                                      a18 == 5 ~ "30000-39000",
                                      a18 == 6 ~ "40000-49000",
                                      a18 == 7 ~ "50000+")),
         income_1yr = as.factor(case_when(a18_rec1 == 0 ~ "$19,000", 
                                      a18_rec1 == 1 ~ "$20,000-$49,000", 
                                      a18_rec1 == 2 ~ "$50,000")), 
         any_util = as.factor(case_when(any_util == 0 ~ "No", any_util == 1 ~ "Yes")), 
         attempted_suicide = as.factor(case_when(g1c == 0 ~ "No", g1c == 1 ~ "Yes")), 
         employment = as.factor(
           case_when(a13 == 1 ~ "Full time", 
                     a13 == 2 ~ "Part time",
                     a13 == 3 ~ "Student", 
                     a13 == 4 ~ "Unemployed",
                     a13 == 5 ~ "Ctrl_envir")), 
         homeless = as.factor(case_when(homeless == 0 ~ "No", homeless == 1 ~ "Yes")),
         hs_grad = as.factor(case_when(hs_grad == 0 ~ "No", hs_grad == 1 ~ "Yes")),
         group = as.factor(case_when(group == 0 ~ "Control", group == 1 ~ "Clinic")),
         # linkstatus = as.factor(case_when(linkstatus == 0 ~ "Did not link to primary care", linkstatus == 1 ~ "Linked to Primary Care")),
         alcohol = as.factor(case_when(alcohol == 0 ~ "Not First Drug", alcohol == 1 ~ "First Drug Alcohol")),
         money_spent_on_alcohol = as.numeric(h16a),
         mh_index = as.numeric(mh),
         num_med_problems = as.numeric(d3),
         num_hospitilizations = as.numeric(d1),
         bothered_by_med = as.factor(case_when(d4 == 0 ~ "Not at all",
                                               d4 == 1 ~ "Slightly",
                                               d4 == 2 ~ "Moderately",
                                               d4 == 3 ~ "Considerably",
                                               d4 == 4 ~ "Extremely")),
         bothered = as.factor(case_when(d4_rec == 0 ~ "No",
                                        d4_rec == 1 ~ "Yes"))) %>% 
  select(group, dayslink, linkstatus, yrs_education, gender, age, alcohol, alcq_30, marriage, employment, income, income_1yr, any_util, attempted_suicide, homeless, hs_grad, money_spent_on_alcohol, mh_index, num_med_problems, num_hospitilizations,bothered_by_med, bothered)
```

```{r}
df %>% 
  mutate(linkstatus = as.factor(case_when(linkstatus == 0 ~ "Did not link to primary care", linkstatus == 1 ~ "Linked to Primary Care"))) %>% 
  select(group, linkstatus, dayslink, income, mh_index, attempted_suicide) %>% 
  ggplot() + geom_point(aes(x=dayslink, y=mh_index, color=linkstatus)) + facet_grid(vars(attempted_suicide), vars(group)) + ylab("Mental Health Index") + xlab("Time") + ggtitle("Mental Health Index Grouped by Attemped Suicide and Study Response")
```

## Cox PH Model

The explanatory variables we want to use are number of medical problems, gender, age, first drug alcohol, treatment, mental health index, and attempted suicide, and control. 

First we will show the survival curves for just the treatment and control groups.
```{r}
care_fit <- survfit(Surv(dayslink, linkstatus) ~ group + employment, data=df)
ggsurvplot(care_fit, conf.int = F, fun="cloglog") + 
  ggtitle("Survival Plot for HELP Treatment")

coxph(Surv(dayslink, linkstatus) ~ group, data=df)
```
The Cox-PH model for just the treatment group tells us that a unit increase in treatment to control groups (0 -> 1) changes the factor of survival by $e^{-1.6415} = 0.1936893$.

```{r}
care_fit <- survfit(Surv(dayslink, linkstatus) ~ group + attempted_suicide, data=df)
ggsurvplot(care_fit, conf.int = T, fun="cloglog") + 
  ggtitle("Survival Plot for HELP Treatment")
coxph(Surv(dayslink, linkstatus) ~ group + attempted_suicide, data=df)
coxph(Surv(dayslink, linkstatus) ~ group, data=df) %>% glance()
base_loglik <- -683.2197	
```

Next we'll compute the drop-in-deviance test to determine if attempted suicide should be included in our model.

$G = 2*(logLik_{bigger model} - logLik_{smaller model})$
```{r}
coxph(Surv(dayslink, linkstatus) ~ group + attempted_suicide, data=df) %>% glance()
G = 2*(-683.0351 - base_loglik)
1-pchisq(G, 1)
```
The change in deviance is 0.3692, $(H_0: \gamma=0)$, so with one degree of freedom the p-value is 0.5434407, which is greater than 0.05. We fail to reject the null hypothesis and do not need type in the model. 

```{r}
coxph(Surv(dayslink, linkstatus) ~ group + mh_index, data=df) %>% glance()
G = 2*(-683.1243 - base_loglik)
1-pchisq(G, 1)
```
The change in deviance is 0.1908, $(H_0: \gamma=0)$, so with one degree of freedom the p-value is 0.6622516, which is greater than 0.05. We fail to reject the null hypothesis and do not need type in the model. 

```{r}
coxph(Surv(dayslink, linkstatus) ~ group + gender, data=df) %>% glance()
G = 2*(-681.652 - base_loglik)
1-pchisq(G, 1)
```
The change in deviance is 3.1354, $(H_0: \gamma=0)$, so with one degree of freedom the p-value is 0.07660959, which is greater than 0.05. Note that the p-value is close to 0.05, which suggests that there could be little evidence. We fail to reject the null hypothesis and do not need type in the model. 

```{r}
coxph(Surv(dayslink, linkstatus) ~ group + alcohol, data=df) %>% glance()
G = 2*(-679.0874 - base_loglik)
1-pchisq(G, 1)
```
The change in deviance is [1] 8.2646. $(H_0: \gamma=0)$, so with one degree of freedom the p-value is 0.004042557, which is less than 0.05. We reject the null hypothesis that $\gamma=0$ in favor of $H_a: \gamma \ne 0$ and should include first drink alcohol in the model. 

```{r}
coxph(Surv(dayslink, linkstatus) ~ group + num_med_problems, data=df) %>% glance()
G = 2*(-681.4424 - base_loglik)
1-pchisq(G, 1)
```
We see that the p-value for the additive model for treatment groups and the number of medical problems is insignificant (0.05938062). Note that the additive model 

```{r}
coxph(Surv(dayslink, linkstatus) ~ group + alcohol, data=df) %>% glance()
base_loglik_alc <- -679.0874 # update base loglikelihood for future drop-in-deviance tests
coxph(Surv(dayslink, linkstatus) ~ group + alcohol + num_med_problems, data=df) %>% glance()
G = 2*(-677.4701 - base_loglik_alc)
1-pchisq(G, 1)
```
The p-value is 0.072. This does not suggest evidence to reject the null hypothesis that $\gamma = 0.$

```{r}
coxph(Surv(dayslink, linkstatus) ~ group + alcohol + gender, data=df) %>% glance()
G = 2*(-678.1924 - base_loglik_alc)
1-pchisq(G, 1)
```

```{r}
coxph(Surv(dayslink, linkstatus) ~ group + alcohol + mh_index, data=df) %>% glance()
G = 2*(-678.8862 - base_loglik_alc)
1-pchisq(G, 1)
```

```{r}
coxph(Surv(dayslink, linkstatus) ~ group + alcohol + attempted_suicide, data=df) %>% glance()
G = 2*(-678.9307 - base_loglik_alc)
1-pchisq(G, 1)
```

**** to use later ****
```{r}
coxph(Surv(dayslink, linkstatus) ~ group + alcohol, data=df) %>% tidy()
```
Explanatory variables we want to use are gender, age, income, employment, homeless, and hs_grad. 
```{r}
coxph(Surv(dayslink, linkstatus) ~ group + age, data=df) %>% glance()
G = 2*(-679.8128 - base_loglik)
1-pchisq(G, 1)
```
0.009045607 < 0.05. We reject the null hypothesis and should include age. 

```{r}
coxph(Surv(dayslink, linkstatus) ~ group + age + alcohol, data=df) %>% glance()
G = 2*(-677.0938 - -679.8128)
1-pchisq(G, 1)
```

```{r}
coxph(Surv(dayslink, linkstatus) ~ group + age + alcohol + employment, data=df) %>% glance()
G = 2*(-675.4454	- -677.0938)
1-pchisq(G, 1)
```

```{r}
coxph(Surv(dayslink, linkstatus) ~ group + age + alcohol + homeless, data=df) %>% glance()
G = 2*(-677.0244 - 	-677.0938	)
1-pchisq(G, 1)
```

```{r}
coxph(Surv(dayslink, linkstatus) ~ group + age + alcohol + income, data=df) %>% glance()
G = 2*(-661.2074 - 	-677.0938)
1-pchisq(G, 1)
```

```{r}
coxph(Surv(dayslink, linkstatus) ~ group + age + alcohol + income + yrs_education, data=df) %>% glance()
G = 2*(-651.3298 - -661.2074)
1-pchisq(G, 1)

coxph(Surv(dayslink, linkstatus) ~ group + age + alcohol + income + yrs_education, data=df) 
```

```{r}
coxph(Surv(dayslink, linkstatus) ~ group + age + alcohol + income + yrs_education + hs_grad, data=df) %>% glance()
G = 2*(-649.1222 - -651.3298)
1-pchisq(G, 1)
```
After running the drop-in-deviance tests, we found age, alcohol, income, years in education, and high school graduate to be significant. Based on Occam's Razor, we are interested in picking a simpler model. Thus, we will proceed with Backward Selection. 

*** 5/8/21: we built a cox ph model using the step-wise method. The model below includes all the variables that passed the drop-in-deviance tests
```{r}
coxph(Surv(dayslink, linkstatus) ~ group + age + alcohol + income + yrs_education + hs_grad, data=df) 

cox.zph(coxph(Surv(dayslink, linkstatus) ~ group + age + alcohol + income + yrs_education + hs_grad, data=df))

care_fit <- survfit(Surv(dayslink, linkstatus) ~ group + alcohol, data=df)
ggsurvplot(care_fit, conf.int = F, fun="cloglog") + 
  ggtitle("Survival Plot for HELP Treatment")

care_fit <- survfit(Surv(dayslink, linkstatus) ~ group + income, data=df)
ggsurvplot(care_fit, conf.int = F, fun="cloglog") + 
  ggtitle("Survival Plot for HELP Treatment")

care_fit <- survfit(Surv(dayslink, linkstatus) ~ group + hs_grad, data=df)
ggsurvplot(care_fit, conf.int = F, fun="cloglog") + 
  ggtitle("Survival Plot for HELP Treatment")
```

```{r}
length(df$yrs_education)
df %>% select(yrs_education) %>% drop_na()
```


