---
title: "Survival_Analysis_Final"
output: pdf_document
---

```{r, warning=FALSE, message=FALSE}
library(mosaic)
library(readr)
library(tidyverse)
library(broom)
library(survival)
library(survminer)
library(praise)
```


```{r, warning=FALSE, message=FALSE}
df <- read_csv("HELPdata.csv", na="*")
```

```{r, message=FALSE}
df <- df %>%
  mutate(yrs_education = as.numeric(a9), 
         gender=a1, 
         alcq_30 = as.numeric(alcq_30), 
         marriage = as.factor(a10), 
         employment = as.factor(a13), 
         income = as.factor(case_when(a18 == 1 ~ "<5000", 
                                      a18 == 2 ~ "5000-10000", 
                                      a18 == 3 ~ "11000-19000",
                                      a18 == 4 ~ "20000-29000",
                                      a18 == 5 ~ "30000-39000",
                                      a18 == 6 ~ "40000-49000",
                                      a18 == 7 ~ "50000+")),
         income_1yr = as.factor(case_when(a18_rec1 == 0 ~ "$19,000", 
                                      a18_rec1 == 1 ~ "$20,000-$49,000", 
                                      a18_rec1 == 2 ~ "$50,000")), 
         any_util = as.factor(case_when(any_util == 0 ~ "No", any_util == 1 ~ "Yes")), 
         attempted_suicide = as.factor(case_when(g1c == 0 ~ "No", g1c == 1 ~ "Yes")), 
         employment = as.factor(
           case_when(a13 == 1 ~ "Full time", 
                     a13 == 2 ~ "Part time",
                     a13 == 3 ~ "Student", 
                     a13 == 4 ~ "Unemployed",
                     a13 == 5 ~ "Ctrl_envir")), 
         homeless = as.factor(case_when(homeless == 0 ~ "No", homeless == 1 ~ "Yes")),
         hs_grad = as.factor(case_when(hs_grad == 0 ~ "No", hs_grad == 1 ~ "Yes")),
         group = as.factor(case_when(group == 0 ~ "Control", group == 1 ~ "Clinic")),
         # linkstatus = as.factor(case_when(linkstatus == 0 ~ "Did not link to primary care", linkstatus == 1 ~ "Linked to Primary Care")),
         alcohol = as.factor(case_when(alcohol == 0 ~ "Not First Drug", alcohol == 1 ~ "First Drug Alcohol")),
         money_spent_on_alcohol = as.numeric(h16a),
         mh_index = as.numeric(mh),
         num_med_problems = as.numeric(d3),
         num_hospitilizations = as.numeric(d1),
         bothered_by_med = as.factor(case_when(d4 == 0 ~ "Not at all",
                                               d4 == 1 ~ "Slightly",
                                               d4 == 2 ~ "Moderately",
                                               d4 == 3 ~ "Considerably",
                                               d4 == 4 ~ "Extremely")),
         bothered = as.factor(case_when(d4_rec == 0 ~ "No",
                                        d4_rec == 1 ~ "Yes"))) %>% 
  select(group, dayslink, linkstatus, yrs_education, gender, age, alcohol, alcq_30, marriage, employment, income, income_1yr, any_util, attempted_suicide, homeless, hs_grad, money_spent_on_alcohol, mh_index, num_med_problems, num_hospitilizations,bothered_by_med, bothered)
```

## Group Members

Michael He and Oliver Chang

## Exploratory Data Analysis

In this exploratory data analysis, we are interested in exploring the relationship years of education, gender, age, alcoholic, and number of drinks in the past 30 days have on whether an individual connected to primary care. 

Below is are statistics of our variables of interest. 
```{r}
summary(df)
```


```{r}
care_fit <- survfit(Surv(dayslink, linkstatus) ~ group, data=df)
ggsurvplot(care_fit, conf.int = T) + 
  ggtitle("Survival Plot for HELP Treatment")
```

```{r}
ggsurvplot_facet(care_fit, df, facet.by = c("gender", "alcohol"), pval = TRUE)
```


```{r, message=FALSE}
df %>% 
  select(age, linkstatus) %>% 
  group_by(age, linkstatus) %>% 
  filter(linkstatus == 1)
  

df %>% 
  select(age, linkstatus) %>% 
  group_by(age, linkstatus) %>% 
  mutate(linkstatus = as.factor(linkstatus)) %>% 
  summarise(count = n()/347) %>% 
  ggplot() + geom_line(aes(x=age, y=count, group=linkstatus, color=linkstatus)) + geom_point(aes(x=age, y=count, group=linkstatus, color=linkstatus)) + ggtitle("Prorption of Link Status Given Age") + ylab("Proportion")

df %>% 
  select(age, linkstatus) %>% 
  group_by(age, linkstatus) %>% 
  mutate(linkstatus = as.factor(linkstatus)) %>% 
  summarise(count = n()) %>% 
  ggplot() + geom_line(aes(x=age, y=count, group=linkstatus, color=linkstatus)) + geom_point(aes(x=age, y=count, group=linkstatus, color=linkstatus)) + ggtitle("Prorption of Link Status Given Age") + ylab("Count")
```

```{r, message=FALSE}
df %>% 
  select(yrs_education, linkstatus) %>% 
  group_by(yrs_education, linkstatus) %>% 
  mutate(linkstatus = as.factor(linkstatus)) %>% 
  summarise(count = n()/347) %>% 
  ggplot() + geom_line(aes(x=yrs_education, y=count, group=linkstatus, color=linkstatus)) + geom_point(aes(x=yrs_education, y=count, group=linkstatus, color=linkstatus))
```

More EDA

```{r}
df
```

```{r}
df %>% 
  select(gender, age) %>% 
  group_by(age, gender) %>% 
  summarise(count = n()) %>% 
  mutate(gender = as.factor(gender)) %>% 
  ggplot() + geom_line(aes(x=age, y=count, color=gender))

df %>% 
  select(gender, alcohol) %>% 
  mutate(gender_str = as.factor(case_when(gender == 1 ~ "Male", gender == 2 ~ "Female"))) %>% 
  mutate(alcohol_str = as.factor(case_when(alcohol == 0 ~ "Not First Drug", alcohol == 1 ~ "First Drug Alcohol"))) %>% 
  ggplot() + geom_bar(aes(x=gender_str, fill=alcohol), position = "fill") + xlab("Gender") + ylab("Proportion")

df %>% 
  select(gender, alcohol) %>% 
  mutate(gender_str = as.factor(case_when(gender == 1 ~ "Male", gender == 2 ~ "Female"))) %>% 
  mutate(alcohol_str = as.factor(case_when(alcohol == 0 ~ "Not First Drug", alcohol == 1 ~ "First Drug Alcohol"))) %>% 
  ggplot() + geom_bar(aes(x=gender_str, fill=alcohol)) + xlab("Gender") + ylab("Number of Observations")
```

Analyze alcq_30
```{r}
df %>% 
  select(alcq_30) %>% 
  group_by(alcq_30) %>% 
  summarise(count = n()) %>% 
  ggplot + geom_histogram(aes(x=alcq_30))
```
Looking at the histogram of the number of drinks past 30 days, we see many outliers after 100 drinks. 

```{r, message=FALSE}
hist(df$yrs_education)

df %>% 
  select(yrs_education, alcohol) %>% 
  mutate(alcohol_str = as.factor(case_when(alcohol == 0 ~ "Not First Drug", alcohol == 1 ~ "First Drug Alcohol")))  %>% 
  group_by(yrs_education, alcohol_str) %>% 
  summarise(count = n()) %>% 
  ggplot + geom_bar(aes(x=yrs_education, y=count), stat="identity") + facet_grid(alcohol_str ~ .)
```

```{r, message=FALSE}
df %>% 
  select(age, alcohol) %>% 
  mutate(alcohol_str = as.factor(case_when(alcohol == 0 ~ "Not First Drug", alcohol == 1 ~ "First Drug Alcohol")))  %>% 
  group_by(age, alcohol_str) %>% 
  summarise(count = n()) %>% 
  ggplot + geom_bar(aes(x=age, y=count), stat="identity") + facet_grid(alcohol_str ~ .)
```

```{r}
df %>% 
  select(age, gender) %>% 
  mutate(gender_str = as.factor(case_when(gender == 1 ~ "Male", gender == 2 ~ "Female"))) %>% 
  group_by(age, gender_str) %>% 
  summarise(count = n()) %>% 
  ggplot + geom_bar(aes(x=age, y=count), stat="identity") + facet_grid(gender_str ~ .)

df %>% 
  select(age, gender) %>% 
  mutate(gender_str = as.factor(case_when(gender == 1 ~ "Male", gender == 2 ~ "Female"))) %>% 
  group_by(age, gender_str) %>% 
  summarise(count = n()) %>% 
  ggplot + geom_line(aes(x=age, y=count,color=gender_str)) + geom_point(aes(x=age, y=count,color=gender_str)) 
```

More EDA
```{r}
df
```

```{r}
df %>% 
  select(group, marriage, linkstatus) %>% 
  ggplot() + geom_bar(aes(x=marriage)) + facet_grid(vars(group), vars(linkstatus))
```

```{r}
df %>% 
  select(group, employment, linkstatus) %>% 
  ggplot() + geom_bar(aes(x=employment)) + facet_grid(vars(group), vars(linkstatus))
```

```{r}
df %>% 
  select(group, hs_grad, linkstatus) %>% 
  drop_na() %>% 
  ggplot() + geom_bar(aes(x=hs_grad)) + facet_grid(vars(group), vars(linkstatus))

df %>% 
  select(group, yrs_education, linkstatus) %>% 
  drop_na() %>% 
  ggplot() + geom_bar(aes(x=yrs_education)) + facet_grid(vars(group), vars(linkstatus))

plot(df$hs_grad)
```

```{r}
par(mfrow=c(2,2))

df %>% 
  select(income, employment) %>% 
  ggplot() + geom_bar(aes(x=income, fill=employment)) + coord_flip()

df %>% 
  select(income_1yr, employment) %>% 
  ggplot() + geom_bar(aes(x=income_1yr, fill=employment)) + coord_flip()
```

```{r}
df %>% 
  select(income, employment, alcohol) %>% 
  ggplot() + geom_bar(aes(x=income, fill=employment)) + facet_grid(vars(alcohol)) + coord_flip()

df %>% 
  select(income, employment, alcohol, group) %>% 
  ggplot() + geom_bar(aes(x=income, fill=employment)) + facet_grid(vars(alcohol), vars(group)) + coord_flip()

df %>% 
  select(income, employment, alcohol, group, linkstatus) %>% 
  ggplot() + geom_bar(aes(x=income, fill=employment)) + facet_grid(vars(alcohol), vars(linkstatus)) + coord_flip()
```

```{r}
df %>% 
  select(group, linkstatus, money_spent_on_alcohol) %>% 
  ggplot()  + geom_point(aes(x=linkstatus, y=money_spent_on_alcohol, color=group), position = position_jitter(w = 0.05, h = .05))
```

```{r}
df %>% 
  select(group, linkstatus, money_spent_on_alcohol, mh_index) %>% 
  ggplot() + geom_point(aes(x=mh_index, y=money_spent_on_alcohol, color=group)) + facet_grid(vars(linkstatus)) 
```

```{r}
df %>% 
  select(group, linkstatus, mh_index, yrs_education) %>% 
  ggplot() + geom_point(aes(x=mh_index, y=yrs_education, color=group)) + facet_grid(vars(linkstatus)) 
```

```{r}
df %>% 
  select(group, linkstatus, dayslink, mh_index, yrs_education) %>% 
  ggplot() + geom_point(aes(x=dayslink, y=mh_index, color=group)) + facet_grid(vars(linkstatus)) 
```
```{r}
df
```

```{r}
df %>% 
  select(group, linkstatus, dayslink, age, yrs_education) %>% 
  ggplot() + geom_point(aes(x=dayslink, y=age, color=group)) + facet_grid(vars(linkstatus)) 
```

```{r}
df %>% 
  select(group, linkstatus, dayslink, alcohol, yrs_education) %>% 
  ggplot() + geom_point(aes(x=dayslink, y=alcohol, color=group)) + facet_grid(vars(linkstatus)) 

df %>% 
  select(group, linkstatus, dayslink, alcq_30, yrs_education) %>% 
  ggplot() + geom_point(aes(x=dayslink, y=alcq_30, color=group)) + facet_grid(vars(linkstatus)) 

df %>% 
  select(group, linkstatus, dayslink, marriage, yrs_education) %>% 
  ggplot() + geom_point(aes(x=dayslink, y=marriage, color=group)) + facet_grid(vars(linkstatus)) 

df %>% 
  select(group, linkstatus, dayslink, employment, yrs_education) %>% 
  ggplot() + geom_point(aes(x=dayslink, y=employment, color=group)) + facet_grid(vars(linkstatus)) 

df %>% 
  select(group, linkstatus, dayslink, income, yrs_education) %>% 
  ggplot() + geom_point(aes(x=dayslink, y=income, color=group)) + facet_grid(vars(linkstatus)) 



df %>% 
  select(group, linkstatus, dayslink, income, mh_index, attempted_suicide, any_util) %>% 
  ggplot() + geom_point(aes(x=dayslink, y=any_util, color=group)) + facet_grid(vars(linkstatus)) 

df %>% 
  select(group, linkstatus, dayslink, yrs_education, mh_index, attempted_suicide, any_util) %>% 
  ggplot() + geom_point(aes(x=dayslink, y=yrs_education, color=group)) + facet_grid(vars(linkstatus)) 

``` 


```{r}
df %>% 
  mutate(linkstatus = as.factor(case_when(linkstatus == 0 ~ "Did not link to primary care", linkstatus == 1 ~ "Linked to Primary Care"))) %>% 
  select(group, linkstatus, dayslink, income, mh_index, attempted_suicide) %>% 
  ggplot() + geom_point(aes(x=dayslink, y=mh_index, color=linkstatus)) + facet_grid(vars(attempted_suicide), vars(group)) + ylab("Mental Health Index") + xlab("Time") + ggtitle("Mental Health Index Grouped by Attemped Suicide and Study Response")
```

```{r}
care_fit <- survfit(Surv(dayslink, linkstatus) ~ group, data=df)

ggsurvplot_facet(care_fit, df, facet.by = c("", "attempted_suicide"), pval = TRUE)
```


```{r}
df %>% 
  mutate(linkstatus = as.factor(case_when(linkstatus == 0 ~ "Did not link to primary care", linkstatus == 1 ~ "Linked to Primary Care"))) %>% 
  select(group, linkstatus, dayslink, num_med_problems, mh_index, attempted_suicide) %>% 
  ggplot() + geom_point(aes(x=dayslink, y=num_med_problems, color=group)) + facet_grid(vars(linkstatus)) + ggtitle("Number of Medical Problems by Days Separated by Success") + ylab("Numner of Medical Problems")

df %>% 
  select(group, linkstatus, dayslink, num_hospitilizations, mh_index, attempted_suicide) %>% 
  ggplot() + geom_point(aes(x=dayslink, y=num_hospitilizations, color=group)) + facet_grid(vars(linkstatus)) 

df %>% 
  select(group, linkstatus, dayslink, bothered_by_med, mh_index, attempted_suicide) %>% 
  ggplot() + geom_point(aes(x=dayslink, y=bothered_by_med, color=group)) + facet_grid(vars(linkstatus)) 

```


## Cox PH Model

The explanatory variables we want to use are number of medical problems, gender, first drug alcohol, treatment, and control. 

First we will show the survival curves for just the treatment and control groups.
```{r}
care_fit <- survfit(Surv(dayslink, linkstatus) ~ group, data=df)
ggsurvplot(care_fit, conf.int = T, fun="cloglog") + 
  ggtitle("Survival Plot for HELP Treatment")

coxph(Surv(dayslink, linkstatus) ~ group, data=df)
```
The Cox-PH model for just the treatment group tells us that a unit increase in treatment to control groups (0 -> 1) changes the factor of survival by $e^{-1.6415} = 0.1936893$. 
```{r}
cox.zph(coxph(Surv(dayslink, linkstatus) ~ group, data=df))
```
Furthermore, the cox.zph shows a p-values of 0.0064 which is less than 0.05. This indicates a violation of proportional hazards.

```{r}
care_fit <- survfit(Surv(dayslink, linkstatus) ~ group + num_med_problems, data=df)
ggsurvplot(care_fit, conf.int = T, fun="cloglog") + 
  ggtitle("Survival Plot for HELP Treatment")

coxph(Surv(dayslink, linkstatus) ~ group  + num_med_problems, data=df)
```

```{r}
care_fit <- survfit(Surv(dayslink, linkstatus) ~ group, data=df)

ggsurvplot(care_fit, conf.int = T) + 
  ggtitle("Survival Plot for HELP Treatment")

coxph(Surv(dayslink, linkstatus) ~ group, data=df)
```

## Someting New

Investigation of the proportional hazards assumption

Michael will investigate the assumptions about proportional hazards. Specifically, it will be about the cox.zph function in R.

Resources to learn include [R Documentation](https://www.rdocumentation.org/packages/survival/versions/3.2-7/topics/cox.zph) for applying the funciton, [UCLA](https://stats.idre.ucla.edu/r/examples/asa/r-applied-survival-analysis-ch-6/) for further explanation, and application of this method in published studies, such as [this one](https://pubmed.ncbi.nlm.nih.gov/10619055/).

The cox.zph function tests every predictor's proportionality. It uses the transform command to create interactions with all the predictors. It is an extension of the coxph function in R and transforms the data using various forms, such as KM or rank. It is directly related to Schoenfeld residuals.

The main challenges in this case are getting the cox.zph function to work and understanding Schoenfeld residuals. The other minor issue will be interpreting the zph graphics, as they look very different from KM curves or previous assignments.

Bootstrapping the survival model (what are the assumptions?  what do you conclude?)

Oliver will take on Bootstrapping the survival model. Bootstraping is a nonparametric technique that resamples the sampled data and samples from the resampled data. It is most useful for building confidence intervals without the assumption of the central limit theory. However, one important assumption is that the observations were randomly sampled. This holds true for this study as the paper mentions 235 observations were randomized into the help clinic. From bootstrapping we hope to estimate the true parameter of the population, which are the coefficients to our proportional hazard model. 

One resource Oliver will use is [censboot](https://stat.ethz.ch/R-manual/R-patched/library/boot/html/censboot.html). Censboot is a function that applies variables bootstrap resampling on survial data. The censboot function references articles that further discuss Bootstraping a survival model. Papers from the [Journal of American Statiscial Association](https://www.jstor.org/stable/2290992?seq=1) and [International Statistics Institute](https://www-jstor-org.ccl.idm.oclc.org/stable/1403683?seq=8#metadata_info_tab_contents) will serve as good references to learn more about Bootstrapping. 

One challenge in applying the bootstrap method in a survival analysis is making sure it works for the type given dataset. In past classes, Oliver has worked with bootstrapping before but never on time and censored dependent data. This challenge warrants further investigation in dataset transformation and the boot package. 


