---
title: "Survival_Analysis_Final"
output: pdf_document
---

```{r, warning=FALSE, message=FALSE}
library(mosaic)
library(readr)
library(tidyverse)
library(broom)
library(survival)
library(survminer)
library(praise)
```


```{r, warning=FALSE, message=FALSE}
df <- read_csv("HELPdata.csv", na="*")
```

```{r}
df <- df %>%
  mutate(yrs_education = as.numeric(a9), gender=a1, alcq_30 = as.numeric(alcq_30)) %>% 
  select(group, dayslink, linkstatus, yrs_education, gender, age, alcohol, alcq_30)
```

## Group Members

Michael He and Oliver Chang

## Exploratory Data Analysis

In this exploratory data analysis, we are interested in exploring the relationship years of education, gender, age, alcoholic, and number of drinks in the past 30 days have on whether an individual connected to primary care. 

Below is are statistics of our variables of interest. 
```{r}
summary(df)
```


```{r}
care_fit <- survfit(Surv(dayslink, linkstatus) ~ group, data=df)
ggsurvplot(care_fit, conf.int = T) + 
  ggtitle("Survival Plot for HELP Treatment")
```

```{r}
ggsurvplot_facet(care_fit, df, facet.by = c("gender", "alcohol"), pval = TRUE)
```


```{r}
df %>% 
  select(age, linkstatus) %>% 
  group_by(age, linkstatus) %>% 
  summarise(count = n()) %>% 
  ggplot() + geom_line(aes(x=age, y=count, group=linkstatus))
```

## Someting New

Investigation of the proportional hazards assumption

Bootstrapping the survival model (what are the assumptions?  what do you conclude?)


